Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2020-05-06 Manuscript/manuscript.Rmd
methods edits (boundary prediction at bp resolution section)

Chromosome-specific datasets used for modeling were composed of two components: the response vector $Y_{i}$, an indicator variable separating between boundary and non-boundary regions, and the predictor-type specific data matrix, $\textbf{X}_{ij}$, where $i \in \{ 1,2, \cdots,N \}$ ($N$ being the number of chromosome-specific bins) and $j \in \{ 1,2,\cdots,p \}$ ($p$ being the number of genomic annotations). <!--Chromosome-specific datasets were concatenated to form a whole genome dataset. The dataset was then split into a 7:3 ratio for training and testing, respectively.[!!!only comes into play with preciseTAD]-->

@@

To assess the impact of class imbalance (CI), defined as the proportion of TAD boundary regions to non-TAD boundary regions, we evaluated three re-sampling techniques including *Random Over-Sampling (ROS)*, *Random Under-Sampling (RUS)*, and *Synthetic Minority Over-Sampling Technique (SMOTE)*. For ROS, the minority class was sampled with replacement to obtain the same number of data points in the majority class. For RUS, the majority class was sampled without replacement to obtain the same number of data points in the minority class. For SMOTE, under-sampling was performed without replacement from the majority class, while over-sampling was performed by creating new synthetic observations using the $k=5$ minority class nearest neighbors [@chawla2002smote] (implemented in the `DMwR` v.0.4.1 R package). We restricted the algorithm to 100% over-sampling and 200% under-sampling, which creates perfectly balanced classes. [??? Of what size?]

@@

Random forest classification models (the `caret` v.6.0 R package [@kuhn2012caret]) were built to compare model performances between combinations of data resolutions, predictor types, and re-sampling techniques. Three-fold cross validation was used to tune the number of annotations, *p*, to consider at each node. The number of trees that were aggregated for each random forest model was set to the default, $n=500$. The performance of each model was compared using Balanced Accuracy (BA), defined as the average of sensitivity and specificity: [??? Correct?]

@@

## TAD boundary prediction at bp resolution

To investigate whether we could alleviate limitations of resolution *r* of Hi-C data that limits the precision of conventional TAD calling algorithms, we applied our optimally built TAD prediction models for predicting TAD boundaries at base pair (bp) resolution. [??? Write how your vector Y and annotation matrices change]. Naturally, when dealing with such fine resolution, there will be spatially close base pairs with similar predictor profiles (i.e. similar overlap counts, distances). To guard against inflating the number of predicted TAD boundaries we developed a technique that combines both density-based spatial clustering (DBSCAN) and partitioning techniques to identify more precise TAD boundaries throughout the linear genome.

>>>>

Chromosome-specific datasets used for modeling were composed of two components: the response vector $Y_{i}$, an indicator variable separating between boundary and non-boundary regions, and the predictor-type specific data matrix, $\textbf{X}_{ij}$, where $i \in \{ 1,2, \cdots,N \}$ ($N$ being the number of chromosome-specific bins) and $j \in \{ 1,2,\cdots,P \}$ ($P$ being the number of genomic annotations). <!--Chromosome-specific datasets were concatenated to form a whole genome dataset. The dataset was then split into a 7:3 ratio for training and testing, respectively.[!!!only comes into play with preciseTAD]-->

@@

To assess the impact of class imbalance (CI), defined as the proportion of boundary regions to non-boundary regions, we evaluated three re-sampling techniques including *Random Over-Sampling (ROS)*, *Random Under-Sampling (RUS)*, and *Synthetic Minority Over-Sampling Technique (SMOTE)*. For ROS, the minority class was sampled with replacement to obtain the same number of data points in the majority class. For RUS, the majority class was sampled without replacement to obtain the same number of data points in the minority class. For SMOTE, under-sampling was performed without replacement from the majority class, while over-sampling was performed by creating new synthetic observations using the $k=5$ minority class nearest neighbors [@chawla2002smote] (implemented in the `DMwR` v.0.4.1 R package). We restricted the algorithm to 100% over-sampling and 200% under-sampling, which creates perfectly balanced classes. [??? Of what size?; !!!not sure what question means]

@@

Random forest classification models (the `caret` v.6.0 R package [@kuhn2012caret]) were built to compare model performances between combinations of data resolutions, predictor types, and re-sampling techniques. Five-fold cross validation was used to tune the number of annotations, *p*, to consider at each node. The number of trees that were aggregated for each random forest model was set to the default, $n=500$. The performance of each model was compared using Balanced Accuracy (BA), defined as the average of sensitivity and specificity: [??? Correct?;!!!yep]

@@

## Boundary prediction at bp resolution

To investigate whether we could alleviate limitations of resolution *r* of Hi-C data that limits the precision of conventional TAD calling algorithms, we applied our optimally built TAD prediction models for predicting TAD boundaries at base pair (bp) resolution. [??? Write how your vector Y and annotation matrices change;!!!added]. At this point, we opted to implement an optimally built whole genome random forest model. That is, a random forest model built on the optimal combination of data resolution (binning), predictor type (feature space), re-sampling technique (class balancing correction), and annotations (most informative genomic elements). That is, $Y$ has length $N$ where $N=n_1+n_2+\cdots+n_{22}$, the total number of bins across each chromosome. The data matrix $\textbf{X}$ has dimensions $N \times p_s$, where $p_s \subset P$, a subset of the most predictive genomic elements.

Naturally, when dealing with such fine resolution, there will be spatially close base pairs with similar predictor profiles (i.e. similar overlap counts, distances). To guard against inflating the number of predicted TAD boundaries we developed a technique that combines both density-based spatial clustering (DBSCAN) and partitioning techniques to identify precise TAD boundaries throughout the linear genome.
